M\"obius Inversion: Construction and Applications

### 2017-04-08 Drafts/Draft.tex
figures

A category is an algebraic structure which generalizes ``objects'' and ``maps'' (or ``morphisms''), with composition. There are several equivalent definitions, all somewhat technical, but the construct itself is quite intuitive once you get used to it.  
\begin{defn}
A \textit{category} $C$ consists of:
\begin{itemize}
\item A collection $C_0$ of \emph{objects};
\item For each $a,b \in C_0$, a collection $C(a,b)$ of \emph{morphisms from $x$ to $y$};
\item For each $a,b,c \in C_0$, a \emph{composition operator}: A function $\circ_{abc} :C(a,b) \times C(b,c) \rightarrow C(a,c)$ (note that we usually omit the subscript, because they can be inferred); and
\item For each $a \in C_0$, an \emph{identity morphism} $1_a \in C(a,a)$,
\end{itemize}
such that the following axioms are satisfied:
\begin{itemize}
\item Composition is associative: If $a,b,c,d \in C_0$ and $f\in C(a,b), g \in C(b,c), h \in C(c,d)$, then $f \circ (g \circ h) = (f \circ g) \circ h$, and
\item Composition respects identities: If $a,b \in C_0, f \in C(a,b)$, then $1_b \circ f = f = f \circ 1_a$.
\end{itemize}
\end{defn}
Note that we sometimes use function notation to write morphisms. So if $f \in C(a,b)$, we might right $f:a \rightarrow b$. We also often write $x \in C$ rather than $x \in C_0$ for an object $x$ of $C$, although technically $C$ is the entire category while $C_0$ is the set of objects.

@@

The following definition is due to \cite{Lein1}.
\begin{defn}\label{cat_euler}
Let $\omega$ be a weighting on a finite category $A$. Then the \emph{Euler-Leinster characteristic}, or just the \emph{Euler characteristic}, of $A$, denoted by $\chi(A)$, is defined by
\[
\chi(A) = \sum_{a \in A} \omega(a)
\]
\end{defn}
Note that even if a category admits several weightings, they all lead to the same Euler characteristic by a simple sum rearrangement argument. Together with \ref{mobius_is_weighting}, this implies the following:
\begin{cor}\label{euler_mu}
If a finite category $A$ has a M\"obius inversion $\mu$, then
\[\chi(A) = \sum_{a \subseteq A} \sum_{b \subseteq A} \mu(a, b).
\]
\end{cor}

\section{Simplicial Complexes}
Calling a new construction an Euler characteristic needs some intuitive justification. We will introduce a structure called a \textit{simplicial complex} to show that calling this value an Euler characteristic is reasonable.

@@

\begin{defn}
A \emph{monoidal category} is an ordinary category $A$, together with
\begin{itemize}
\item A \emph{monoidal product} $\otimes : A \times A \rightarrow A$;
\item A \emph{unit object} $u \in A$;
\end{itemize}
such that
\begin{itemize}
\item For each ordered triple of objects $x,y,z \in A$, $(x \otimes y) \otimes z \cong x \otimes (y \otimes z)$
\item For each $x \in A$, $u \otimes x \cong x \cong x \otimes u$,
\end{itemize} 
and all the isomorphisms commute appropriately.
\end{defn}
There are a variety of commutation requirements which we are not considering in detail here. For a full account, see \cite{Kelly1}.

@@

We can consider some examples of norms.
\begin{examp}
Let $D$ be the monoidal category of non-negative real numbers defined in example \ref{cat_real}. Then, for all $t \in \Rr$, the function $|\cdot|:D \rightarrow \Rr$ defined by 
\[
|x| = e^{-tx}
\]
is a norm on $D$.

We will explore this norm in detail shortly.
\end{examp}

@@

We are interested in a notion of how close together things are - and it will need to measure closeness rather than distance, in the sense that it will increase for objects which are closer together. The reason for this will become clear soon. We do this by defining $\zeta$ appropriately: If $C$ is a generalized metric space, then $\zeta(a,b) = e^{-C(a,b)}$.

@@

There is another way of looking at essentially the same property. The reason the addition (+) operator appears in the second axiom in the generalized metric space definition is because addition is what is called the \textit{monoidal product} in the enriched category. The equivalent in an ordinary category is the tensor product, $\otimes$. In terms of number of morphisms, $\otimes$ behaves like multiplication. We want a fact about addition to give us a fact about multiplication. Exponentiating is the most natural way to translate multiplication into addition, so we can apply the axiom.
\section{Product Spaces and Approximations}
\section{Examples}
\end{chapter}

% % % Extra stuff/deleted portions
\iffalse
\iffalse
We can replace the vectors $f,g$ with matrices to get the following generalization, which :
\begin{thm}[Two Variable M\"obius Inversion]\label{mob_inv_old}
Let $X$ be a lower finite poset.  Let $f,g:X \times X \rightarrow \Rr$ be given. Let $\zeta:X \times X \rightarrow \Rr$ be defined as usual, and let $\mu = \zeta^{-1}$. Then:
\begin{align*}
f = g \zeta &\iff g = f \mu;\\
f = \zeta g &\iff g = \mu f.
\end{align*}
Equivalently,
\begin{align*}
\forall x,y \in X, f(x,y) = \sum_{z \in X}g(x,z)\zeta(z,y) &\iff \forall x,y \in X, g(x,y) = \sum_{z \in X}f(x,z)\mu(z,y);\\
\forall x,y \in X, f(x,y) = \sum_{z \in X}\zeta(x,z)g(z,y) &\iff \forall x,y \in X, g(x,y) = \sum_{z \in X}\mu(x,z)f(z,y).
\end{align*}
\begin{proof}
Again, the result is trivial when we treat $f,g,\zeta,\mu$ as matrices.
\end{proof}
\end{thm}
\fi

The \textbf{identity function} $\delta$ is defined by 
\[\delta(a,b) = \begin{cases} 1 & a = b \\ 0 & else. \end{cases}
\]
\fi
\end{document}

>>>>

A category is an algebraic structure which generalizes ``objects'' and ``maps'' (or ``morphisms''), with composition. There are several equivalent definitions, all somewhat technical, but the construct itself is quite intuitive once you get used to it.  
\begin{defn}
A \textit{category} $C$ consists of:
\begin{itemize}
\item A collection $C_0$ of \emph{objects};
\item For each $a,b \in C_0$, a set $C(a,b)$ of \emph{morphisms from $x$ to $y$}, sometimes called the \emph{hom-set} of $(a,b)$;
\item For each $a,b,c \in C_0$, a \emph{composition operator}: A function $\circ_{abc} :C(a,b) \times C(b,c) \rightarrow C(a,c)$ (note that we usually omit the subscript, because they can be inferred); and
\item For each $a \in C_0$, an \emph{identity morphism} $1_a \in C(a,a)$,
\end{itemize}
such that the following axioms are satisfied:
\begin{itemize}
\item Composition is associative: If $a,b,c,d \in C_0$ and $f\in C(a,b), g \in C(b,c), h \in C(c,d)$, then $f \circ (g \circ h) = (f \circ g) \circ h$, and
\item Composition respects identities: If $a,b \in C_0, f \in C(a,b)$, then $1_b \circ f = f = f \circ 1_a$.
\end{itemize}
\end{defn}
Note that we sometimes use function notation to write morphisms. So if $f \in C(a,b)$, we might right $f:a \rightarrow b$. We also often write $x \in C$ rather than $x \in C_0$ for an object $x$ of $C$, although technically $C$ is the entire category while $C_0$ is the set of objects.

@@

The following definition is due to \cite{Lein1}.
\begin{defn}\label{cat_euler}
Let $\omega$ be a weighting on a finite category $A$. Then the \emph{Euler-Leinster characteristic}, or just the \emph{Euler characteristic}, of $A$, denoted by $\chi(A)$, is defined by
\[
\chi(A) = \sum_{a \in A} \omega(a)
\]
\end{defn}
Note that even if a category admits several weightings, they all lead to the same Euler characteristic by a simple sum rearrangement argument. Together with \ref{mobius_is_weighting}, this implies the following:
\begin{cor}\label{euler_mu}
If a finite category $A$ has a M\"obius inversion $\mu$, then
\[\chi(A) = \sum_{a \subseteq A} \sum_{b \subseteq A} \mu(a, b).
\]
\end{cor}
%TODO: Think of category w/ no euler char
\section{Simplicial Complexes}
Calling a new construction an Euler characteristic needs some intuitive justification. We will introduce a structure called a \textit{simplicial complex} to show that calling this value an Euler characteristic is reasonable.

@@

\begin{defn}\label{def_monoid}
A \emph{monoidal category} is an ordinary category $A$, together with
\begin{itemize}
\item A \emph{monoidal product} $\otimes : A \times A \rightarrow A$;
\item A \emph{unit object} $u \in A$;
\end{itemize}
such that
\begin{itemize}
\item For each ordered triple of objects $x,y,z \in A$, $(x \otimes y) \otimes z \cong x \otimes (y \otimes z)$
\item For each $x \in A$, $u \otimes x \cong x \cong x \otimes u$,
\item The function $\otimes$ is \emph{functorial}. This essentially means that if $x_1, x_2, y_1, y_2 \in M, f \in C(x_1, x_2), g \in C(y_1, y_2)$, then there exists a corresponding morphism in $C(x_1 \otimes y_1, x_2 \otimes y_2)$. We call that morphism $f \otimes g$, and require that it behave as we would expect (e.g. if $f,g$ are both isomorphisms, $f \otimes g$ also is).
\end{itemize} 
\end{defn}
There are a variety of commutation requirements which we are not considering in detail here. The property of being functorial is also more involved. For a full account, see \cite{Kelly1}.

@@

We can consider some examples of norms.
\begin{examp}\label{exp_norm}
Let $D$ be the monoidal category of non-negative real numbers defined in example \ref{cat_real}. Then, for all $t \in \Rr$, the function $|\cdot|:D \rightarrow \Rr$ defined by 
\[
|x| = e^{-tx}
\]
is a norm on $D$. We can see that it satisfies the requirements:
\[|0| = 1
\]
\[|x + y| = e^{-tx + ty} = |x||y|.
\]
We will explore this norm in detail shortly.
\end{examp}

@@

%TODO: Sports crowd analogy

So we would like a norm on $D$, the monoidal category of real numbers, which decreases as its argument get large. Recall that in example \ref{exp_norm}, we observed that the function $|x|=e^{-tx}$ is a norm on $D$. For any positive $t$, this is a decreasing function. In fact, we wills see that if we use this family of norms, we can view $t$ as a ``zooming'' parameter, which allows us to consider the data at different scales.

We have a special name for the Euler characteristic of a generalized metric space which results from this particular norm:
\begin{defn}
Let $M$ be a generalized metric space. Let $|\cdot|_t:D \rightarrow \Rr$ be defined by
\[
|x|_t = e^{-tx}.
\]
If it exists, let $\chi_t(M)$ be the Euler characteristic resulting from $|\cdot|$. Then $\chi_t(M)$ is called the \emph{$t$-magnitude} (or just \emph{magnitude} if $t=1$) of $M$, denoted $\magn_t(M)$ or just $\magn(M)$.
\end{defn}
%TODO: Small examples w/ graphs
\section{Product Spaces and Approximations}

Consider the following set of points $A$ in two-dimensional space:

\[
\includegraphics{../Figures/4x3.pdf}
\]

Let $M$ be this set:
\[
\includegraphics{../Figures/3.pdf}
\]

and $N$ this set:
\[
\includegraphics{../Figures/4.pdf}
\]

The set $A$ is what you would get if you replaced each point in $M$ with the set $N$. Suppose we choose a scale, and somehow count the number of clusters in $N$ and in $M$.

If the scale is very small -- if we ``zoom in'' so that points must be very close together before they are considered to be in the same cluster -- then every point will be in its own cluster. So $N$ will have four clusters, $M$ three, and $A$ will have 12.

If we choose a middle scale, then it might be that the points in $N$ are clustered together, but $M$ remains discrete. Then $N$ has one cluster, $M$ has three, and $A$ has four, as each copy of $N$ forms a cluster.

Finally, if our scale is large enough that $M$ is grouped into a single cluster, all three sets of points will have a single cluster.

In all three cases, the number of clusters in $A$ is the product of the number of clusters in $M$ and in $N$. It appears that this remains the case as we change our scale.

$A$ is an approximation of what is called a \emph{tensor product category}. In this section, we define the construction formally, and explore how the Euler-Leinster characteristic behaves when two categories are combined in this way. In particular, we would like for the magnitude of a product space to be the product of the magnitudes of the original spaces - this is how the number of clusters seems like it should behave when we look at examples like $A$ above.

Not all enriched categories can be used to construct product spaces. We first must define a restriction on the definition of a monoidal category.

\begin{defn}
Let $M$ be a monoidal category with monoidal product $\otimes$. Suppose that for all $a,b \in M, a \otimes b \cong b \otimes a$. Then $M$ is a \emph{symmetric monoidal category}.
\end{defn}

We add this restriction to allow us to prove the following lemma:

\begin{lemma}
Let $M$ be a symmetric monoidal category. Let $P,Q,R,S,T,U$ in $M$ be given. Suppose that we have morphisms
\begin{align*}
f&:(P \otimes Q) \rightarrow T\\
g&:(R \otimes S) \rightarrow U.
\end{align*}
Then, if $f \otimes g$ is the morphism required by the functoriality requirement of definition \ref{def_monoid}, we have
\[
(f \otimes g):(P \otimes Q) \otimes (R \otimes S) \rightarrow T \otimes U.
\]
\end{lemma}
\begin{proof}\label{prod_morph}
Using the symmetry of $\otimes$, we can rearrange:
\[
(P \otimes Q) \otimes (R \otimes S) \cong (P \otimes R) \otimes (Q \otimes S).
\]
Then definition \ref{def_monoid} tells us that the desired morphism exists.
\end{proof}
This extra requirement is necessary for the product of two enriched categories to exist. In particular, the composition maps do not necessarily exist if the monoidal category is not symmetric. We can now define that:

\begin{defn}
Let $A,B$ be categories both enriched over a symmetric monoidal category $M$. Fix a norm on $M$. Then the \emph{tensor product} of $A$ and $B$, denoted $A \times B$, is the category enriched over $M$ defined as follows:
\begin{itemize}
\item The set of objects, $(A \times B)_0$, is $\{(a,b):a \in A_0, b \in  B_0\}$
\item For $(a,b), (c,d) \in A_0, C((a,b), (c,d)) = C(a,b) \otimes C(c,d)$.
\item The unit object is $u_A \otimes u_B$, where $u_A, u_B$ are the units of $A,B$ respectively
\item The composition and identity morphisms are defined using Lemma \ref{prod_morph} (the details are not important for this discussion, but they are straightforward).
\end{itemize}
\begin{rmk}
The tensor product of two categories enriched over $M$ is indeed a category enriched over $M$. There are many requirements to check, and we do not go through all of them here.
\end{rmk}
\end{defn}

We consider some examples of simple product categories. Note that \catname{set} is a symmetric monoidal category, so ordinary categories are eligible to be used as factors of product categories. 
\begin{examp}
Let $A$ be the following category:
\[
\includegraphics{../Figures/cat_2_directed.pdf}
\]
Recall that we omit the identity morphisms when we draw categories. So $A$ has two points and three morphisms. Then $A \times A$ looks like this:
\[
\includegraphics{../Figures/cat_2_prod.pdf}
\]
To see why, recall that the monoidal product on \catname{set} is the direct product. So consider, for example, $(a,a)$ and $(a,b)$. There is one morphism from $a$ to $a$ and one from $a$ to $b$. So the direct product of the hom-sets also has one element, which is why there is one arrow from $(a,a)$ to $(a,b)$. On the other hand, because there is no morphism $b$ to $a$, there is also no morphism from $(a,b)$ to $(a,a)$. 
\end{examp}
\begin{examp}\label{met_prod_ex}
We will be focusing on products of generalized metric spaces. So we consider an example of that. Let $A, B$ be the following:
\[
\includegraphics{../Figures/gen_met_small.pdf}
\]
So each space has two points, with $A(a_1, a_2) = 1, B(b_1, b_2) = 2$. Then the tensor product category $A \times B$ looks like this:
\[
\includegraphics[width=.4\paperwidth]{../Figures/gen_met_small_prod}
\]
It is impossible to draw this product space to scale because of the way distances behave.

We can observe the same behavior here as in the initial example. Suppose we used some clustering scheme where points within a threshold distance $t$ of each other were put into the same cluster. This is not how magnitude works, of course, but it is easier to visualize. Then we have three cases:
\begin{description}
\item[$t < 1$:] $A$ has two clusters, $B$ has two clusters, and $A \times B$ has four clusters.
\item[$1 \leq t < 2$:] $A$ has one cluster, $B$ has two clusters, and $A \times B$ has two clusters.
\item[$t \geq 2$:] $A, B$, and $A \times B$ each have just one cluster (note that $(a_1, b_2)$ and $(a_2, b_1)$ are in the same cluster in $A \times B$ even though their distance is 3, because both are in the same cluster as $(a_1, b_1)$).
\end{description}
So in all cases, the number of clusters in $A \times B$ is the product of the numbers of clusters in $A$ and $B$. We will see that this isn't a coincidence.
\end{examp}

Magnitude is obtained using $\zeta$, the similarity matrix. So we are interested in how the similarity matrix behaves under the tensor product. It turns out that the similarity matrix of a product category is obtained using the matrix \emph{Kronecker product}, not coincidentally (but confusingly) written using $\otimes$. 
\begin{defn}
Let $A$ be an $m \times n$ matrix and $B$ be a $q \times r$ matrix. Then the \emph{Kronecker product} of $A$ and $B$, denoted $A \otimes B$, is the $mq \times nr$ matrix defined by the following block matrix:
\[
A \otimes B = 
\begin{bmatrix}
A_{11}B & A_{12}B & \cdots & A_{1n}B \\
A_{21}B & A_{22}B & \cdots & A_{2n}B \\
\vdots & \vdots & \ddots & \vdots \\
A_{m1}B & A_{m2}B & \cdots & A_{mn}B
\end{bmatrix}.
\]
\end{defn}
We show that this definition makes sense by connecting it immediately to the tensor product of categories.

\begin{lemma}\label{tensor_kron}
Let $A, B$ be categories both enriched over a monoidal category $M$. Fix some norm on $M$. Let $\zeta_A, \zeta_B$ be the similarity matrices of $A,B$ under the chosen norm. If $a_1,\dots,a_m$ and $b_1,\dots,b_n$ are the orders chosen for the objects of $A$ and $B$ when constructing $\zeta_A, \zeta_B$, then fix the order \[a_1b_1, a_1b_2, \dots, a_1b_n, a_2b_1, \dots, a_mb_n\] for $A \times B$. Then
\[
\zeta_{A \times B} = \zeta_A \otimes \zeta_B.
\]
\end{lemma}
\begin{proof}
%TODO: Prove
\end{proof}
\begin{thm}
Let $A, B$ be categories both enriched over a monoidal category $M$ . Fix some norm on $M$. Then, if $\chi(A), \chi(B)$ both exist,
\[
\chi(A \times B) = \chi(A)\chi(B).
\]
\end{thm}
\begin{proof}
Let $\zeta_A, \zeta_B$ be the similarity matrices of $A,B$. Then, by theorem \ref{tensor_kron}, the similarity matrix of $A \times B$ is $\zeta_A \otimes \zeta_B$.

Because both Euler characteristics exist, there must exist a valid weighting on each category. Let $\omega_A, \omega_B$ be weightings on $A, B$ respectively. Let $1^{|A|}, 1^{|B|}$ be the ones vectors of lengths $|A|$ and $|B|$. Then, using facts about the Kronecker product, we have
\begin{align*}
1^{|A|}&= \zeta_A\omega_A \\
1^{|B|}&= \zeta_B\omega_A \\
1^{|A||B|} &= \zeta_A\omega_A \otimes \zeta_B\omega_B\\
&= (\zeta_A \otimes \zeta_B)(\omega_A \otimes \omega_B)
\end{align*}
So the vector $\omega_A \otimes \omega_B$ is a weighting on $A \otimes B$. We have:
\begin{align*}
\chi(A \times B) &= \sum_{\alpha \in A \otimes B}(\omega_A \otimes \omega_B)(\alpha) \\
&= \sum_{a \in A, b \in B} \omega_A(a)\omega_B(b) \\
&= \left(\sum_{a \in A}\omega_A(a)\right)\left(\sum_{b \in B}\omega_B(b)\right) \\
&= \chi(A)\chi(B).
\end{align*}
\end{proof}
\section{Examples}
%TODO: Necessary to be its own section?
\end{chapter}

% % % Extra stuff/deleted portions
\iffalse

We can replace the vectors $f,g$ with matrices to get the following generalization, which :
\begin{thm}[Two Variable M\"obius Inversion]\label{mob_inv_old}
Let $X$ be a lower finite poset.  Let $f,g:X \times X \rightarrow \Rr$ be given. Let $\zeta:X \times X \rightarrow \Rr$ be defined as usual, and let $\mu = \zeta^{-1}$. Then:
\begin{align*}
f = g \zeta &\iff g = f \mu;\\
f = \zeta g &\iff g = \mu f.
\end{align*}
Equivalently,
\begin{align*}
\forall x,y \in X, f(x,y) = \sum_{z \in X}g(x,z)\zeta(z,y) &\iff \forall x,y \in X, g(x,y) = \sum_{z \in X}f(x,z)\mu(z,y);\\
\forall x,y \in X, f(x,y) = \sum_{z \in X}\zeta(x,z)g(z,y) &\iff \forall x,y \in X, g(x,y) = \sum_{z \in X}\mu(x,z)f(z,y).
\end{align*}
\begin{proof}
Again, the result is trivial when we treat $f,g,\zeta,\mu$ as matrices.
\end{proof}
\end{thm}

The \textbf{identity function} $\delta$ is defined by 
\[\delta(a,b) = \begin{cases} 1 & a = b \\ 0 & else. \end{cases}
\]

@@

There is another way of looking at essentially the same property. The reason the addition (+) operator appears in the second axiom in the generalized metric space definition is because addition is what is called the \textit{monoidal product} in the enriched category. The equivalent in an ordinary category is the tensor product, $\otimes$. In terms of number of morphisms, $\otimes$ behaves like multiplication. We want a fact about addition to give us a fact about multiplication. Exponentiating is the most natural way to translate multiplication into addition, so we can apply the axiom.
\fi
\end{document}
